---
title: "Getting Started with pgscatalogcolors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pgscatalogcolors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r load}
library(pgscatalogcolors)
library(ggplot2)
```

## What ancestry acronyms are available?

The package maps **19 ancestry categories** defined by the
[PGS Catalog ancestry documentation](https://www.pgscatalog.org/docs/ancestry/).
Call `pgscatalog_ancestry_info()` to see the full lookup table:

```{r info}
pgscatalog_ancestry_info()
```

The columns are:

| Column | Description |
|--------|-------------|
| `acronym` | The **exact string** your data column must contain |
| `ancestry` | Full PGS Catalog ancestry label |
| `pop_group` | Broader population group abbreviation (1000G/HGDP-style) |
| `color` | Hex color assigned to that acronym |

## What does my data frame need?

Your data frame needs **one column whose values are ancestry acronyms** — the
exact strings from the `acronym` column above (case-sensitive). The column can
have any name; you point ggplot2 to it with `aes(color = <your_column>)`.

```{r data-requirements}
# Minimal working example
df <- data.frame(
  sample_id  = paste0("s", 1:6),
  pgs_score  = c(0.42, -0.11, 0.87, 0.03, -0.55, 0.29),
  ancestry   = c("EUR", "AFR", "EAS", "SAS", "HIS", "MID")
  #              ^-- must match acronyms exactly
)
df
```

The `ancestry` column values must be **character or factor** — numeric codes
will not match. The column does **not** need to contain all 19 acronyms; only
the ones present in your data are plotted.

## Basic scatter plot

```{r scatter}
set.seed(1)
n   <- 200
grp <- c("EUR", "AFR", "EAS", "SAS", "HIS", "MID")

df <- data.frame(
  PC1      = rnorm(n),
  PC2      = rnorm(n),
  ancestry = sample(grp, n, replace = TRUE,
                    prob = c(0.40, 0.15, 0.15, 0.12, 0.10, 0.08))
)

ggplot(df, aes(PC1, PC2, color = ancestry)) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_pgscatalog() +
  labs(title = "PCA by ancestry", color = "Ancestry") +
  theme_classic()
```

## Bar chart (fill aesthetic)

```{r bar}
counts <- data.frame(
  ancestry = c("EUR", "AFR", "EAS", "SAS", "HIS", "MID", "MIE", "MEE"),
  n        = c(512, 198, 175, 143, 97, 61, 44, 18)
)

ggplot(counts, aes(reorder(ancestry, -n), n, fill = ancestry)) +
  geom_col() +
  scale_fill_pgscatalog() +
  labs(x = "Ancestry", y = "Count", title = "Sample counts by ancestry") +
  theme_classic() +
  theme(legend.position = "none")
```

## Using `scale_color_manual()` directly

If you need more control (e.g., custom legend labels), use
`pgscatalog_scale_values()` to get only the colors for groups in your data:

```{r manual}
my_groups <- c("EUR", "AFR", "EAS")

ggplot(df[df$ancestry %in% my_groups, ], aes(PC1, PC2, color = ancestry)) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = pgscatalog_scale_values(my_groups),
    labels = c(AFR = "African", EAS = "East Asian", EUR = "European")
  ) +
  theme_classic()
```

## Checking colors programmatically

```{r check}
# Look up specific colors by acronym
pgscatalog_cols("EUR", "AFR", "EAS")

# Check whether your acronyms are all recognized before plotting
my_data_groups <- c("EUR", "AFR", "XYZ")   # "XYZ" is not a valid acronym
pgscatalog_scale_values(my_data_groups)     # warns about unknown acronyms
```

## What if my data uses full ancestry names instead of acronyms?

Join against `pgscatalog_ancestry_info()` to add the acronym column:

```{r join}
raw <- data.frame(
  sample_id = 1:3,
  ancestry_long = c("European", "African unspecified", "East Asian")
)

info <- pgscatalog_ancestry_info()
merged <- merge(raw, info[, c("ancestry", "acronym")],
                by.x = "ancestry_long", by.y = "ancestry")
merged
```
