---
title: "Getting Started with pgscatalogcolors"
author: "Brooke N. Wolford (ORCID: 0000-0003-3153-1552)"
date: "pgscatalogcolors `r packageVersion('pgscatalogcolors')` — `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pgscatalogcolors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r load}
library(pgscatalogcolors)
library(ggplot2)
```

## What ancestry acronyms are available?

The package maps **19 ancestry categories** defined by the
[PGS Catalog ancestry documentation](https://www.pgscatalog.org/docs/ancestry/).
Call `pgscatalog_ancestry_info()` to see the full lookup table:

```{r info}
pgscatalog_ancestry_info()
```

The columns are:

| Column | Description |
|--------|-------------|
| `acronym` | The **exact string** your data column must contain |
| `ancestry` | Full PGS Catalog ancestry label |
| `pop_group` | Broader population group abbreviation (1000G/HGDP-style) |
| `color` | Hex color assigned to that acronym |

## What does my data frame need?

Your data frame needs **one column whose values are ancestry acronyms** — the
exact strings from the `acronym` column above (case-sensitive). The column can
have any name; you point ggplot2 to it with `aes(color = <your_column>)`.

```{r data-requirements}
# Minimal working example
df <- data.frame(
  sample_id  = paste0("s", 1:6),
  pgs_score  = c(0.42, -0.11, 0.87, 0.03, -0.55, 0.29),
  ancestry   = c("EUR", "AFR", "EAS", "SAS", "HIS", "MID")
  #              ^-- must match acronyms exactly
)
df
```

The `ancestry` column values must be **character or factor** — numeric codes
will not match. The column does **not** need to contain all 19 acronyms; only
the ones present in your data are plotted.

## PGS odds ratio forest plot

A common use case is plotting the odds ratio (OR) of a polygenic score across
ancestry groups, with ancestry on the y-axis and the OR on the x-axis. Each
point is coloured by ancestry using the PGS Catalog palette.

```{r forest-plot, fig.height = 4}
set.seed(42)
groups <- c("EUR", "AFR", "EAS", "SAS", "HIS", "MID", "MIE", "MEE")

forest_df <- data.frame(
  ancestry = factor(groups, levels = rev(groups)),
  OR       = c(1.85, 1.42, 1.63, 1.71, 1.38, 1.55, 1.60, 1.44),
  lower    = c(1.72, 1.28, 1.49, 1.57, 1.21, 1.35, 1.41, 1.22),
  upper    = c(1.99, 1.58, 1.79, 1.87, 1.58, 1.78, 1.81, 1.70)
)

ggplot(forest_df, aes(x = OR, y = ancestry, color = ancestry)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.25, linewidth = 0.7) +
  geom_point(size = 3.5) +
  scale_color_pgscatalog() +
  scale_x_continuous(breaks = seq(1, 2, 0.25)) +
  labs(
    x     = "Odds Ratio (95% CI)",
    y     = "Ancestry",
    title = "PGS performance by ancestry group"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

## Basic scatter plot

```{r scatter}
set.seed(1)
n   <- 200
grp <- c("EUR", "AFR", "EAS", "SAS", "HIS", "MID")

df <- data.frame(
  PC1      = rnorm(n),
  PC2      = rnorm(n),
  ancestry = sample(grp, n, replace = TRUE,
                    prob = c(0.40, 0.15, 0.15, 0.12, 0.10, 0.08))
)

ggplot(df, aes(PC1, PC2, color = ancestry)) +
  geom_point(size = 2.5, alpha = 0.85) +
  scale_color_pgscatalog() +
  labs(title = "PCA by ancestry", color = "Ancestry") +
  theme_classic()
```

## Bar chart (fill aesthetic)

```{r bar}
counts <- data.frame(
  ancestry = c("EUR", "AFR", "EAS", "SAS", "HIS", "MID", "MIE", "MEE"),
  n        = c(512, 198, 175, 143, 97, 61, 44, 18)
)

ggplot(counts, aes(reorder(ancestry, -n), n, fill = ancestry)) +
  geom_col() +
  scale_fill_pgscatalog() +
  labs(x = "Ancestry", y = "Count", title = "Sample counts by ancestry") +
  theme_classic() +
  theme(legend.position = "none")
```

## Using `scale_color_manual()` directly

If you need more control (e.g., custom legend labels), use
`pgscatalog_scale_values()` to get only the colors for groups in your data:

```{r manual}
my_groups <- c("EUR", "AFR", "EAS")

ggplot(df[df$ancestry %in% my_groups, ], aes(PC1, PC2, color = ancestry)) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = pgscatalog_scale_values(my_groups),
    labels = c(AFR = "African", EAS = "East Asian", EUR = "European")
  ) +
  theme_classic()
```

## Checking colors programmatically

```{r check}
# Look up specific colors by acronym
pgscatalog_cols("EUR", "AFR", "EAS")

# Check whether your acronyms are all recognized before plotting
my_data_groups <- c("EUR", "AFR", "XYZ")   # "XYZ" is not a valid acronym
pgscatalog_scale_values(my_data_groups)     # warns about unknown acronyms
```

## Cumulative risk curves by PGS quintile

When multiple curves belong to the same ancestry group — for example,
PGS quintile-stratified cumulative risk curves — use `pgscatalog_tints()` to
generate perceptually ordered color variants derived from that ancestry's base
color. Use `facet_wrap()` to display multiple ancestries side by side, with each
panel using that ancestry's own tint ramp.

```{r tints-basic}
# Five EUR tint variants, light to dark
pgscatalog_tints("EUR", n = 5)
```

```{r tints-data, echo=FALSE}
ancestries <- c("EUR", "AFR", "EAS", "SAS")
quintiles  <- paste0("Q", 1:5)

line_cols <- unlist(lapply(ancestries, function(anc)
  pgscatalog_tints(anc, n = 5,
                   labels    = paste0(anc, "_", quintiles),
                   direction = "light_to_dark")))

q_scales <- c(0.4, 0.7, 1.1, 1.7, 2.6)
age_seq  <- seq(40, 80, length.out = 50)
b        <- 0.07

df_tints <- do.call(rbind, lapply(ancestries, function(anc) {
  do.call(rbind, lapply(seq_along(quintiles), function(qi) {
    a    <- 0.00012 * q_scales[qi]
    risk <- 1 - exp(-(a / b) * (exp(b * (age_seq - 40)) - 1))
    data.frame(age = age_seq, risk = risk,
               ancestry = anc, group = paste0(anc, "_", quintiles[qi]))
  }))
}))
df_tints$group <- factor(df_tints$group,
                         levels = unlist(lapply(ancestries,
                                                function(a) paste0(a, "_", quintiles))))
```

```{r tints-curves, fig.width = 9, fig.height = 3.5}
ggplot(df_tints, aes(age, risk, colour = group, group = group)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = line_cols,
                     breaks = paste0("EUR_", quintiles),
                     labels = quintiles, name = "PGS\nQuintile") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ ancestry, nrow = 1) +
  labs(x = "Age (years)", y = "Cumulative risk",
       title = "Cumulative risk by PGS quintile") +
  theme_classic()
```

## Cumulative risk curves with 95% CI ribbons

For plots with per-quintile 95% CI ribbons, build the colour and fill scales
manually so both aesthetics share the same tint ramp. Place `geom_ribbon()`
*before* `geom_line()` so ribbons sit behind lines.

```{r tints-ci-narrow-data, echo=FALSE}
cv_narrow <- 0.20
df_ci <- do.call(rbind, lapply(ancestries, function(anc) {
  do.call(rbind, lapply(seq_along(quintiles), function(qi) {
    a    <- 0.00012 * q_scales[qi]
    risk <- 1 - exp(-(a / b) * (exp(b * (age_seq - 40)) - 1))
    se   <- cv_narrow * risk
    data.frame(age = age_seq, risk = risk,
               lower = pmax(risk - se, 0), upper = risk + se,
               ancestry = anc, group = paste0(anc, "_", quintiles[qi]))
  }))
}))
df_ci$group <- factor(df_ci$group,
                      levels = unlist(lapply(ancestries,
                                             function(a) paste0(a, "_", quintiles))))
```

```{r tints-ribbon-narrow, fig.width = 9, fig.height = 3.5}
ggplot(df_ci, aes(age, risk, colour = group, fill = group, group = group)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.3, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = line_cols,
                     breaks = paste0("EUR_", quintiles),
                     labels = quintiles, name = "PGS\nQuintile") +
  scale_fill_manual(values  = line_cols,
                    breaks  = paste0("EUR_", quintiles),
                    labels  = quintiles, name = "PGS\nQuintile") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ ancestry, nrow = 1) +
  labs(x = "Age (years)", y = "Cumulative risk",
       title = "Non-overlapping CIs") +
  theme_classic()
```

When CIs overlap heavily, draw the darkest ribbons first by reversing the
factor levels on a separate ribbon data frame. This ensures lighter (lower-risk)
ribbons are painted on top, preserving readability.

> **When does this approach break down?** Reversing the draw order helps when
> adjacent bands *partially* overlap — the lighter ribbons remain visible on
> top. However, if confidence intervals are so wide that they completely engulf
> one another (e.g., Q1's band fully covers Q5's at all ages), no draw order
> can recover individual bands. In that situation, consider: (1) displaying
> fewer quintiles, (2) adding direct line labels instead of a ribbon fill, or
> (3) splitting groups into separate facets so each panel's CI bands are
> readable on their own.
>
> **Number of strata:** `pgscatalog_tints()` issues a warning when `n > 8`
> (adjacent variants may be hard to distinguish in dense legends) and a
> stronger warning when `n > 10` (adjacent variants likely to be confused).
> These warnings fire automatically whenever you call `pgscatalog_tints()` or
> any of the scale wrappers with a large `n`.

```{r tints-ci-wide-data, echo=FALSE}
cv_wide <- 0.55
df_ci2 <- do.call(rbind, lapply(ancestries, function(anc) {
  do.call(rbind, lapply(seq_along(quintiles), function(qi) {
    a    <- 0.00012 * q_scales[qi]
    risk <- 1 - exp(-(a / b) * (exp(b * (age_seq - 40)) - 1))
    se   <- cv_wide * risk
    data.frame(age = age_seq, risk = risk,
               lower = pmax(risk - se, 0), upper = risk + se,
               ancestry = anc, group = paste0(anc, "_", quintiles[qi]))
  }))
}))
df_ci2$group <- factor(df_ci2$group,
                       levels = unlist(lapply(ancestries,
                                              function(a) paste0(a, "_", quintiles))))
df_rib2 <- df_ci2
df_rib2$group <- factor(df_rib2$group,
                        levels = unlist(lapply(ancestries,
                                               function(a) paste0(a, "_", rev(quintiles)))))
```

```{r tints-ribbon-wide, fig.width = 9, fig.height = 3.5}
ggplot() +
  geom_ribbon(data = df_rib2,
              aes(x = age, ymin = lower, ymax = upper,
                  fill = group, group = group),
              alpha = 0.3, colour = NA) +
  geom_line(data = df_ci2,
            aes(x = age, y = risk, colour = group, group = group),
            linewidth = 0.9) +
  scale_color_manual(values = line_cols,
                     breaks = paste0("EUR_", quintiles),
                     labels = quintiles, name = "PGS\nQuintile") +
  scale_fill_manual(values  = line_cols,
                    breaks  = paste0("EUR_", quintiles),
                    labels  = quintiles, name = "PGS\nQuintile") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ ancestry, nrow = 1) +
  labs(x = "Age (years)", y = "Cumulative risk",
       title = "Overlapping CIs - dark drawn first, light on top") +
  theme_classic()
```

## Color vision deficiency

The PGS Catalog palette was not originally designed with color vision deficiency
(CVD) in mind. Use `pgscatalog_cvd()` to simulate how the palette appears under
deuteranopia, protanopia, and tritanopia, and to identify which ancestry pairs
fall below a perceptual distance threshold (CIEDE2000 delta-E).

```{r cvd, message = FALSE}
library(colorspace)
library(farver)

# Returns a data frame of flagged pairs, sorted by delta-E ascending
cvd_df <- pgscatalog_cvd(threshold = 20)
print(cvd_df[, c("cvd_type", "acronym1", "acronym2", "delta_e")])
```

Delta-E thresholds to interpret the results:

| Delta-E | Interpretation |
|---------|----------------|
| < 2.3 | Perceptually identical under that CVD type |
| 2.3 – 10 | Difficult to distinguish, especially for small plot elements |
| 10 – 20 | Potentially confusable; consider a supplementary encoding |

### Known problematic pairs

The most severe conflicts in the palette are:

**Deuteranopia** (red-green; affects ~8% of males — the most clinically
relevant type):

- **ASN / HIS** — delta-E **1.7** (perceptually identical; brown and red
  become almost the same olive-brown)
- **EUR / SAS** — delta-E **4.0** (blue and purple converge strongly)
- **MID / MIE** — delta-E **5.4**
- **NON / MIE** — delta-E **7.9**

**Protanopia** (red-green):

- **ASN / HIS** — delta-E **5.5**
- **EAS / MEE** — delta-E **6.0**
- **MID / NON** — delta-E **6.1**

**Tritanopia** (blue-yellow; rare, < 0.01%):

- **MIE / MEE** — delta-E **8.9**
- **ASN / HIS** — delta-E **7.7**

**Normal vision** (included for completeness):

- **ABO / NON** — delta-E **9.8** (both grey by design)
- **ASN / HIS** — delta-E **12.5**

### Recommendations

When a plot includes ancestry groups that form a problematic pair, supplement
colour with a second visual encoding:

```{r cvd-shape, fig.height = 3.5}
set.seed(7)
groups <- c("EUR", "SAS", "EAS", "AFR")   # EUR/SAS are a deuteranopia risk pair

df_cvd <- data.frame(
  PC1      = rnorm(120),
  PC2      = rnorm(120),
  ancestry = rep(groups, each = 30)
)

# Add shape mapping so EUR and SAS are distinguishable without colour
ggplot(df_cvd, aes(PC1, PC2, color = ancestry, shape = ancestry)) +
  geom_point(size = 2.5) +
  scale_color_pgscatalog() +
  scale_shape_manual(values = c(EUR = 16, SAS = 17, EAS = 15, AFR = 18)) +
  labs(title = "Colour + shape encoding for CVD safety") +
  theme_classic()
```

## What if my labels have typos or inconsistent formatting?

Use `pgscatalog_match()` to fuzzy-match a column of messy ancestry labels to
their PGS Catalog acronyms. Matching tries exact acronym, then exact full name,
then edit-distance fuzzy matching — in that order.

```{r fuzzy-basic}
messy <- c("EUR", "eur", "european", "East Asian", "east_asian", "EAST ASIAN")
pgscatalog_match(messy, quiet = TRUE)
```

Fuzzy matches always produce a warning so you can verify them. The `max_dist`
argument (default `3`) controls how many edits are allowed — lower values are
stricter:

```{r fuzzy-warn}
# Default max_dist = 3
pgscatalog_match(c("Europeen", "sas", "Sub-Saharan African"))

# Stricter: only accept 1-edit differences
pgscatalog_match(c("Europeen", "sas", "Sub-Saharan African"), max_dist = 1)

# More lenient: accept up to 5 edits
pgscatalog_match(c("Afrikan", "sth_asian"), max_dist = 5)
```

Labels with no match within `max_dist` return `NA` with a warning:

```{r fuzzy-na}
pgscatalog_match(c("EUR", "martian", "unknown_ancestry"))
```

The typical workflow is to call `pgscatalog_match()` once to add a clean acronym
column, then pass that column to the ggplot2 scales:

```{r fuzzy-workflow}
df_messy <- data.frame(
  sample_id     = paste0("s", 1:6),
  pgs_score     = c(0.42, -0.11, 0.87, 0.03, -0.55, 0.29),
  ancestry_raw  = c("european", "African American or Afro-Caribbean",
                    "east_asian", "South Asian", "HIS", "mid")
)

df_messy$ancestry <- pgscatalog_match(df_messy$ancestry_raw, quiet = TRUE)
df_messy[, c("ancestry_raw", "ancestry")]
```

## What if my data uses full ancestry names instead of acronyms?

For exact full names, `pgscatalog_match()` handles this directly. For cases
where you want to inspect or customise the mapping, you can also join against
`pgscatalog_ancestry_info()` manually:

```{r join}
raw <- data.frame(
  sample_id = 1:3,
  ancestry_long = c("European", "African unspecified", "East Asian")
)

info <- pgscatalog_ancestry_info()
merged <- merge(raw, info[, c("ancestry", "acronym")],
                by.x = "ancestry_long", by.y = "ancestry")
merged
```

## Citing this package

If you use `pgscatalogcolors` in a publication, please cite it as follows:

```{r citation}
citation("pgscatalogcolors")
```

You can also retrieve the citation in BibTeX format:

```{r citation-bib, comment = ""}
toBibtex(citation("pgscatalogcolors"))
```
